---
title: "Filtering joins"
author: Andrew Gard
format: html
embed-resources: true
---

```{r}
#| label: setup
#| message: false

library(tidyverse)

msleep_animals <- data.frame(name = c("Cheetah",
                                      "Dog",
                                      "Cow",
                                      "Goat"),
                             genus = c("Acinonyx",
                                       "Canis",
                                       "Bos",
                                       "Capri"))
msleep_animals

msleep_vore <- data.frame(name = c("Cheetah", 
                                   "Cow",
                                   "Goat",
                                   "Brown bat"),
                          vore = c("carni",
                                   "herbi",
                                   "herbi",
                                   "insecti"))
msleep_vore
```

Often our data is split between two data sets, as is the case with `msleep_animals` and `msleep_vore`, both based on information from the `ggplot2::msleep` data set. You can find the code used to create these sets (ie, this Quarto document) on my website <https://equitableequations.com/>, which you should totally check out anyway.

Filtering joins remove observations from one data set based on whether or not they have matches in a second data set. Like `dplyr::filter`, they do not add or modify columns. 

## The `semi_join` function

`semi_join` retains only observations with a match in the secondary set. For instance, the following code keep only rows in `msleep_animals` that also appear in `msleep_vore`. This means omitting the dog. 

```{r}
msleep_animals |> 
  semi_join(msleep_vore,
            by = "name")
```

The `by` argument specifies how observations should be matched up (here, using the `name` column). This works exactly same way as it does for `left_join` and other mutating joins, including the default behavior when `by` is omitted.

```{r}
msleep_animals |> 
  semi_join(msleep_vore)
```

When the key column has different names in the two sets, use `join_by`. The next chunk creates a new set, `msleep_animals2`, in which the `name` column name is capitalized, then performs the same `semi_join` as before

```{r}
msleep_animals2 <- msleep_animals |> 
  rename("Name" = "name") 

msleep_animals2 |> 
  semi_join(msleep_vore,
            by = join_by("Name" == "name"))
```

Within `join_by`, we specify the names of the corresponding columns, starting with the one used in the primary data set, separated by a double equal sign. 

## The `anti_join` function

To keep only observations that do **not** appear in a second data set, use `anti_join`. For instance, the next code chunk retains only the dog from `msleep_animals`.

```{r}
msleep_animals |> 
  anti_join(msleep_vore,
            by = "name")
```

When taken together, `semi_join` an `anti_join` include all rows from the original data set. 

```{r}
bind_rows(semi_join(msleep_animals,
                    msleep_vore,
                    by = "name"),
          anti_join(msleep_animals,
                    msleep_vore,
                    by = "name"))
```

The output of the above chunk is just `msleep_animals` with the rows re-ordered. 

## The `inner_join` function

`inner_join` adds columns from a secondary data set to the target set, like `left_join`, while keeping only observations that also appear in the secondary set, like `semi_join`.

```{r}
msleep_animals |> 
  inner_join(msleep_vore, 
             by = "name")
```

While still considered a mutating join, `inner_join` performs a filter as well. 





