---
title: "Grouped summaries"
format: html
embed-resources: true
---

```{r}
#| label: setup
#| message: false

library(tidyverse)
```

Today we're summarizing the classic `iris` data set quantitatively. Learn more about the origins of this set with `?iris`. 

The `summarize` command gives a one-row data frame with specified summary statistics about the target set.

```{r}
iris |> 
  summarize(petal_length_avg = mean(Petal.Length),
            petal_length_sd = sd(Petal.Length))
```

The `summarize` command gets more interesting when we use `.by` to obtain results at the group level.

```{r}
iris |> 
  summarize(petal_length_avg = mean(Petal.Length),
            petal_length_sd = sd(Petal.Length),
            .by = Species)
```

This is equivalent to the older (but still useful) syntax

```{r}
#| eval: false

iris |> 
  group_by(Species) |> 
  summarize(petal_length_avg = mean(Petal.Length),
            petal_length_sd = sd(Petal.Length))
```

Here `group_by` just adds a grouping attribute to the set, which you can imagine like color-coding rows of your data set.

```{r}
iris |> 
  group_by(Species) 
```

One advantage of the `group_by` syntax is that it allows grouping on the fly, whereas `.by` only accepts column names. 

```{r}
iris |> 
  group_by(long_sepal = (Sepal.Length >= 6)) |> 
  summarize(petal_length_avg = mean(Petal.Length),
            petal_length_sd = sd(Petal.Length))
```

This is equivalent to:

```{r}
#| eval: false

iris |> 
  mutate(long_sepal = (Sepal.Length >= 6)) |> 
  summarize(petal_length_avg = mean(Petal.Length),
            petal_length_sd = sd(Petal.Length),
            .by = long_sepal)
```

You can group by more than one variable at a time:

```{r}
iris |> 
  mutate(long_sepal = (Sepal.Length >= 6)) |> 
  summarize(petal_length_avg = mean(Petal.Length),
            petal_length_sd = sd(Petal.Length),
            .by = c(Species, 
                    long_sepal))
```

The result is always ungrouped set.

