---
title: "Introducing the pipe"
author: Andrew Gard
format: html
embed-resources: true
---

```{r}
#| message: false

library(tidyverse)
```

Often in R, we need to perform multiple operations on the same data set. Code can get messy fast. 

```{r}
head(mutate(select(filter(penguins, species == "Gentoo"), species, body_mass), body_mass_lb = body_mass * .002204623), n = 10)
```

There are lots of ways to make this chunk clearer, but by far the best option is the pipe operator `|>`, which passes the object on its left to the function on its right as the first argument. 

```{r}
penguins |>  
  filter(species == "Gentoo") |>  
  select(species, body_mass) |>  
  mutate(body_mass_lb = body_mass * .002204623) |>  
  head(n = 10)
```

Code written with the pipe can often be read as a sentence. "Take the penguins set, filter it to retain only gentoos, then select the species and body_mass variables, then add a column showing body mass in pounds, and then display the first 10 rows of the set."

Most major `tidyverse` functions (and many non-`tidyverse` ones) take a data set as the first argument and return a data set at the end, so the pipe is useful in most function calls.

The pipe was added to base R in version 4.1.0. Prior to that, the `magrittr` pipe, `%>%`, was universal. It still loads with `tidyverse`. The following code is equivalent to the chunks above.

```{r}
#| eval: false

penguins %>%  
  filter(species == "Gentoo") %>%  
  select(species, body_mass) %>%  
  mutate(body_mass_lb = body_mass * .002204623) %>%  
  head(n = 10)
```

In general, you should start a new line after a pipe, writing one function per line. For other best-practices, see the tidyverse style guide at <https://style.tidyverse.org/pipes.html>.

It's possible (though less common) to pipe into other arguments than the first.

```{r}
penguins |> 
  lm(body_mass ~ bill_len + species,
     data = _)
```



