---
title: "Mutating joins"
format: html
embed-resources: true
---

```{r}
#| label: setup
#| message: false

library(tidyverse)

msleep_animals <- data.frame(name = c("Cheetah", 
                               "Dog", 
                               "Cow",
                               "Goat"),
                      genus = c("Acinonyx",
                                "Canis",
                                "Bos",
                                "Capri"))

msleep_vore <- data.frame(name = c("Cheetah", 
                                   "Cow",
                                   "Goat",
                                   "Brown bat"),
                          vore = c("carni",
                                   "herbi",
                                   "herbi",
                                   "insecti"))
```

Often our data is split between two data sets, as in the data sets `msleep_animals` and `msleep_vore`, both based on information from the `ggplot2::msleep` data set. You can find the code used to create these sets (ie, this Quarto document) on my website <https://equitableequations.com/>, which you should totally check out anyway.

```{r}
msleep_animals
msleep_vore
```

It's natural to view the second set, `msleep_vore`, as including more information about observations in the first. We'd like to add this information to `msleep_animals`. While it is possible to do this with `mutate`, a simpler approach is to use `left_join`. 

In addition to the two sets, we also specify the **key** column (or columns), the ones which identify observations in the two sets. The output of `left_join` adds all the non-key columns from the second set to the first while keeping observations consistent.

```{r}
msleep_animals |> 
  left_join(msleep_vore,
            by = "name")
```

The order in which the data sets are specified in `left_join` can be very important. If we reverse the roles of the sets above, for instance, we get a different result.

```{r}
msleep_vore |>  
  left_join(msleep_animals, 
            by = "name")
```

## More on key columns

If we omit the `by` argument from `left_join`, R will use all columns that appear in both sets as keys. This is a so-called *natural join*. 

```{r}
msleep_animals |>  
  left_join(msleep_vore)
```

I generally advise against taking advantage of this. It's too easy to overlook subtleties in the structure and names of columns in different data sets.

In all of our examples so far, a single column with a common name sufficed to identify observations in both sets, but this frequently isn't the case. For instance, suppose the creator of our `msleep_vore` set chose to capitalize the title of the `name` column:

```{r}
msleep_vore2 <- msleep_vore |>  
  rename(Name = name) 
msleep_vore2
```

To combine these sets, a more complex `by` argument is needed within `left_join`.

```{r}
msleep_animals |>  
  left_join(msleep_vore2, 
            by = join_by(name == Name))
```

The argument of `join_by` is a vector, so if more than one column is needed to identify observations, you just need to add additional elements inside it. 

## More complicated relationships

Sometimes you'll have more than one observation in the second set for some of the corresponding observations in the first, or vice-versa. This is the case with the `babynames` and `applicants` sets in the `babynames` package.

```{r}
library(babynames)

glimpse(applicants) # one row per year and gender
glimpse(babynames) # many rows per year and gender
```

Notice how `left_join` handles each of these situations.

```{r}
applicants |> left_join(babynames,
                        by = join_by(year, sex))

babynames |> left_join(applicants,
                       by = join_by(year, sex))
```

## Other mutating joins

For most purposes, `left_join` is all you'll need, but sometimes you may have use for more specialized variations. 

- `right_join` keeps only observations that exist in the second data set, adding columns from the first.

- `full_join` keeps all observations that exist in either set (or both).

- `inner_join` keeps **only** observations in the first set with a match in the second, performing a `filter` as well as a `mutate`.

A second data set can be used to `filter` the primary set in several different ways. These so-called *filtering joins* will be covered next in this course.


