---
title: "Tidying data: pivot_wider"
format: html
embed-resources: true
---

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(GLMsData)
data(blocks)
```


## Tidy data

All the major functions in the `ggplot2` and `dplyr` packages (among many others) assume that your data is in a **tidy** format, meaning that:

-   each data set holds one and only one type of observation

-   each column holds one and only one variable

-   each row holds one and only one observation

When a set doesn't have these properties, we begin by reshaping it. Many of the best tools for doing this are included in the `tidyr` package, including `pivot_wider`, which is useful when variable names are found in the cells of a spreadsheet

## The `blocks` dataset 

The `GLMsData::blocks` set includes data from a block-stacking experiment in 1931 in which 25 children stacked 2 sorts of blocks in two trials each. Currently, each row represents one child in one trial with one shape, which is logical for many purposes.

```{r}
glimpse(blocks) 
```

Sometimes, however, we might need to consider an observation to consist of a single child (25 observations total) or a single child on a single trial (50 observations total). In these cases, we expect separate columns for each trial (both examples) and possibly for each shape as well (first example only).

## How to `pivot_wider`

To pivot a too-long data set like this, we need to specify 3 things:

- the variables (columns) that uniquely identify observations.

- The column(s) whose cells currently contain variable names

- The column(s) whose cells currently contain the corresponsing variable values.

The following chunk reshapes `blocks` so that each child represents a single observation. This means that `Trial` and `Shape` are no longer values of variables that help identify observations. Instead, their values should be incorporated into column names, like so:

```{r}
blocks_kids <- blocks |> 
  pivot_wider(id_cols = Child, 
              names_from = c(Trial, Shape),
              values_from = c(Number, Time))
glimpse(blocks_kids)
```

## Another `blocks` pivot

The following chunk reshapes `blocks` so that observations are defined to be a single child stacking a single shape of block, so `Trial` is no longer an identifying variable. 

```{r}
blocks_kidshape <- blocks |> 
  pivot_wider(id_cols = c(Child, Shape), 
              names_from = Trial,
              values_from = c(Number, Time))
glimpse(blocks_kidshape)
```

There are many optional arguments that may be helpful in obtaining more descriptive column names, including `names_sep`.

```{r}
blocks_kidshape <- blocks |> 
  pivot_wider(id_cols = c(Child, Shape), 
              names_from = Trial,
              values_from = c(Number, Time),
              names_sep = "_Trial_")
glimpse(blocks_kidshape)
```

## A dumbbell plot

As an application of this, we might want to create a dumbbell plot comparing the number of cylindrical blocks stacked in the first vs second trial. 

```{r}
blocks_kidshape |> 
  filter(Shape == "Cylinder") |> 
  ggplot() + 
  geom_segment(aes(x = Number_Trial_1, 
                   y = Child, 
                   xend = Number_Trial_2,
                   yend = Child)) + 
  geom_point(aes(x = Number_Trial_1, 
                 y = Child,
                 color = "Trial 1"),
             size = 2.5) +
  geom_point(aes(x = Number_Trial_2, 
                 y = Child,
                 color = "Trial 2"),
             size = 2.5) +
  labs(x = "Number stacked",
       color = NULL,
       title = "Stack 'em to the heavens") +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal()
```




