---
title: "Single-variable plots in R"
format: html
embed-resources: true
---

```{r}
#| label: setup
#| message: false

library(tidyverse) # loads ggplot2
```

In this vid, we'll use the `diamonds` data set to practice single-variable plots using the `ggplot` function.

## One categorical variable

The go-to plot for a single categorical variable is the bar chart. 

```{r}
ggplot(diamonds,
       aes(x = cut)) +
  geom_bar() 
```

Adding color:

```{r}
ggplot(diamonds,
       aes(x = cut)) +
  geom_bar(fill = "limegreen",
           color = "darkgreen") +
  theme_minimal()
```

Reverse the orientation:

```{r}
ggplot(diamonds,
       aes(y = cut)) +
  geom_bar(fill = "limegreen",
           color = "darkgreen") +
  theme_minimal()
```

Order the factor levels by frequency:

```{r}
ggplot(diamonds,
       aes(x = fct_infreq(cut))) +
  geom_bar(fill = "limegreen",
           color = "darkgreen") +
  theme_minimal()
```

Fix labels:

```{r}
ggplot(diamonds,
       aes(x = fct_infreq(cut))) +
  geom_bar(fill = "limegreen",
           color = "darkgreen") +
  labs(x = "Cut",
       y = "Count",
       title = "Fair cut diamonds are rare") +
  theme_minimal()
```

See my vid on pie charts [here](https://youtu.be/ueLozxgEnDc). Use them with extreme care!

## One quantitative variable

The go-to plot for a single quantitative variable is the histogram. 

```{r}
ggplot(diamonds,
       aes(x = price)) +
  geom_histogram()
```

Change the number of bins:

```{r}
ggplot(diamonds,
       aes(x = price)) +
  geom_histogram(binwidth = 1000,
                 boundary = 0) + # or bins 
  theme_minimal()
```

Change to a frequency polygon:

```{r}
ggplot(diamonds,
       aes(x = price)) +
  geom_freqpoly(binwidth = 1000, 
                 boundary = 0) +
  theme_minimal()
```

A density plot:

```{r}
theme_set(theme_minimal())

ggplot(diamonds,
       aes(x = price)) +
  geom_density(fill = "#5798A6",
               linewidth = .2)
```

For very small data sets, you can also consider `geom_dotplot`. 

